---
title: "AnalysisR"
author: "Carlos Arauz"
date: "10/11/2021"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(hrbrthemes)
library(writexl)
library(readxl)
library(lubridate)
library(skimr)
library(forcats)
library(corrplot)
library(data.table)
library(quantmod)
library(hash)
library(gapminder)
library(GGally)

```
#Loading databases

##Loading first data set

This dataset contains Global Food Prices data from the World Food Programme 
covering foods such as maize, rice, beans, fish, and sugar for 76 countries 
and some 1,500 markets.It is updated weekly but contains to a large extent 
monthly data. The data goes back as far as 1992 for a few countries, 
although many countries started reporting from 2003 or thereafter.
   
Source: https://data.humdata.org/dataset/wfp-food-prices

```{r}
global.food.prices.database.wfp.1 <- read.csv("~/Desktop/Project/Food Prices/Databases/R-Data/global-food-prices-database-wfp-1.csv")
head(global.food.prices.database.wfp.1)
```





##Loading second data set
   
   This data set containts agriculture percent of GDP by year by country
   
```{r}
  agriculture_percent_of_gdp <- read.csv("~/Desktop/Project/Food Prices/Databases/R-Data/agriculture_percent_of_gdp.csv")
      head(agriculture_percent_of_gdp)
```

Source: https://data.worldbank.org/indicator/NV.AGR.TOTL.ZS

##Loading third Data Set 

containts Population by year by country 
Source: https://data.worldbank.org/indicator/SP.POP.TOTL

```{r}
library(readxl)
 population <- read_excel("~/Desktop/Project/Food Prices/Databases/R-Data/population.xls")
 population2 <- read_excel("~/Desktop/Project/Food Prices/Databases/R-Data/population2.xls")
```


##Loading 4th data set

This database containts Gdp Per Capita by year by country
 
 #Source:https://data.worldbank.org/indicator/NY.GDP.PCAP.CD
 
 
```{r}
gdp_per_cap <- read.csv("~/Desktop/Project/Food Prices/Databases/R-Data/gdp_per_cap.csv")
head(gdp_per_cap)
```

##Loading 5th data set

data set with PPP Dollars

https://data.worldbank.org/indicator/PA.NUS.PPP

```{r}

ppp <- read.csv("~/Desktop/Project/Food Prices/Databases/R-Data/ppp_world.csv")
head (ppp)

```
#Loading 6th dataset


Source: http://data.un.org/Data.aspx?d=WDI&f=Indicator_Code%3aEP.PMP.SGAS.CD


```{r}
oil_prices<- read.csv("~/Desktop/Project/Food Prices/Databases/R-Data/UNdata_Export_20211204_012639355.csv")
oil_prices<-oil_prices %>% select(1,2,3)
oil_prices<-oil_prices %>% rename("oil_price" = "Value","country"= "Country.or.Area","year"="Year")
oil_prices<-transform(oil_prices, year = as.double(year))
```





#Cleaning data to join databases
 

##database1: 
 
getting rid of rows that will not be used and renaming rows name and adding
purchasing power parity to database

```{r}
ppp<-ppp %>% select(1,5:25)

colnames(ppp)[1] <- "country"

 ppp1 <- gather(ppp,year,purchasingpp,-country)
 
ppp1$year <- strtoi(substring(ppp1$year,2))

ppp1[is.na(ppp1)] <- 0
  
ppp1<-ppp1 %>% filter(purchasingpp>0)


```

##database1: 
 
getting rid of rows that will not be used and renaming rows name and adding
purchasing power parity to database

```{r}
gf<- global.food.prices.database.wfp.1 %>% rename("country" = "adm0_name", "year"= "mp_year","city" = "adm1_name","product"="cm_name")

gf<-inner_join(gf,ppp1,by=c("country","year"))

 gf<-  gf %>% select(2,8,10,12,14,16,17,19) %>%  filter(mp_price<10000,pt_name=="Retail")
 
```


summarizing data by products with same country and year.

```{r}
gf<-gf %>% 
    group_by(product,country,year,um_name,purchasingpp,cur_name) %>% 
    summarize(price=sum(mp_price),
              count=n()
              )
```


```{r}
gf<-transform(gf, price_average = price / count)
```

Only using Kilograms as it is the measurement that appears the most. 

```{r}
gf<-gf %>% filter(um_name=="KG",)
```


changing all the results from diffrent types of currency to usd

```{r}
unique.currency = unique(gf$cur_name)
from <- unique.currency
to <- rep("USD",length(from))
conversion <- getQuote(paste0(from, to, "=X"))
summary(is.na(conversion$Last))
 
current.price <-  conversion$Last
h= hash()
h[from]=current.price

conversion1<-conversion %>% select(2)

conversion2<-tibble::rownames_to_column(conversion1, "currency")


conversion3<-conversion2 %>% separate(currency, c("cur_name", "value"), "USD=X")

conversion4<-conversion3 %>% select(1,3)


gf2<-inner_join(gf,conversion4,by="cur_name")

gf2<-transform(gf2,USD_price = price_average*Last)

gf2<-gf2 %>% select(1:3,5,8,11)


gf2<-gf2 %>% separate(product, c("main_product", "description"), " ", extra = "merge")


head(gf2)

```


#database2: 

Cleaning database editing, so thtat years appear as a column

```{r}
 colnames(agriculture_percent_of_gdp)[1] <- "country"
 
 agri_per <- gather(agriculture_percent_of_gdp,year,agr_gdp_per,-country)
 
agri_per$year <- strtoi(substring(agri_per$year,2))


agri_per[is.na(agri_per)] <- 0
  
agri_per2<-agri_per %>% filter(agr_gdp_per>0)

```


#Database3:

#Joining all years together and editing, so that years apper as a column

```{r}
population01<-population %>% select(1,5:50)

colnames(population01)[1] <- "country"

population1 <- gather(population01,year,population,-country)


colnames(population2)[1]<-"country"

population02<- gather(population2,year,population,-country)


population03<-full_join(population1,population02)


population03 <- population03 %>%  
   mutate(year =   suppressWarnings(as.numeric(year)))


```



#Database4:

Renaming columns to join tables


```{r}
gdp_per_cap01<-gdp_per_cap %>% select(1,5:66)

colnames(gdp_per_cap01)[1] <- "country"

gdp_per_cap1 <- gather(gdp_per_cap01,year,gdp_per_cap,-country)

gdp_per_cap1$year <- strtoi(substring(gdp_per_cap1$year,2))


gdp_per_cap1[is.na(gdp_per_cap1)] <- 0

gdp_per_cap_02<-gdp_per_cap1 %>% filter(gdp_per_cap>1)


```


# Joining tables

```{r}
data1<-inner_join(gf2,agri_per2,by=c("country","year"))

data2<-inner_join(population03,gdp_per_cap_02,by=c("country","year"))

database<-inner_join(data1,data2,by=c("country","year"))

gapminder1<-gapminder %>% select(1,2)

gapminder2 <- unique( gapminder1 )

database<-inner_join(database,gapminder2,by=("country"))
   

```


 This is the database we have now
 
 
 
```{r}
head(database)

```
 
 Now we have our final database where we have information of the product, country,
 continent, usd price, countries agriculture as percent to GDP, GDP Per capita,
 and population.the data range is from 1992-2018, but mostly from 2011-2018.
The mean price of our items was 0.89 USD with a high sd of 1.32






































Lets make a new variable to have the average price by country

```{r}

database01<-database %>% 
  group_by(main_product,year) %>% 
  summarize(count=n(),
            total_price=sum(USD_price))

database01<-transform(database01,ave_price = total_price/count)

head(database01)

```


Lets make a variable that have the sum of the top 5 products with most data points to calculate if they have had 
a similiar behavior of price fluctuation 
```{r}
targetd<-c("Rice","Maize","Wheat",
"Sorghum","Beans")
database02<-filter(database01, main_product %in% targetd)
database02<-transform(database02,ave_price = total_price/count)

database02<-database02 %>% 
  group_by(year) %>% 
  summarize(ave_price=sum(ave_price))


ggplot(database02, aes(x=year, y=ave_price)) + 
   geom_point(size=3) 
```


It seems the ave_price of the combination of the top 5 products increase by year.





##Lets do a regression analysis to see if variable price can be explained by 
variable time

```{r}
food.reg<-lm(ave_price ~ year, data=database02) 
plot(ave_price ~ year, data=database02) 
abline(food.reg$coef)
```

#Basic diagnosis
```{r}
summary(food.reg)
```


##checking residuals

```{r}
 library(broom)
tidy(food.reg)
augment(food.reg)
augment(food.reg) %>%  ggplot(aes(x = year, y = .resid))+
   geom_point() + geom_hline(aes(yintercept = 0))
```


Histogram and Normality plot



```{r}
par(mfrow=c(1,2)) 
hist(food.reg$resi) 
qqnorm(food.reg$resi) 
qqline(food.reg$resi)
```

The histogram is skewed th the right. There are gaps, but nothing to be worried about

The dots in the Q-Q plot appears close to the reference line.


## Test of normality

```{r}
shapiro.test(food.reg$residuals)
```

Since P-value is 0.74 we accept the null hypothesis and believe the residuals
are normally distributed

##Test of linearity

```{r}
food.lof<-lm(ave_price ~ as.factor (year), data=database02)
anova(food.reg, food.lof)
```


At a P-value of 0.90 we accept the null hypothesis and conclude the model
fits the data

```{r}
summary(food.reg)
```


Regression Formula= Y= -42.08+ 0.021X
As time increases price tends to increase lets analyze more profoundly why.




```{r}
summary(database)
```



Organizing countries by there GDP-Per-Capita to see if a countries GDP
affects food prices

 
```{r}
databasenew<-database %>% 
  mutate(class_gdp = ifelse(between(gdp_per_cap, 113.5, 639), "low_class",
                                   ifelse(between(gdp_per_cap, 639, 981.9), "low_medium_class",
                                          ifelse(between(gdp_per_cap, 981.9, 2687.5),"high_medium_class",
                                                 ifelse(between(gdp_per_cap, 2687.5, 12614.8), "high_class", "error")))))
databasenew %>% count(class_gdp,continent)
```

```{r}

database01new<-databasenew %>% 
  group_by(main_product,year,class_gdp) %>% 
  summarize(count=n(),
            total_price=sum(USD_price))

database01new<-transform(database01new,ave_price = total_price/count)

head(database01new)


```

```{r}

targetd<-c("Rice","Maize","Wheat")

database02new<-filter(database01new, main_product %in% targetd)
database02new<-transform(database02new,ave_price = total_price/count)

database02new<-database02new %>% 
  group_by(year,class_gdp) %>% 
  summarize(conteo=n(),ave_price=sum(ave_price))

database02new<-database02new %>% filter(conteo==3)

ggplot(database02new, aes(x=year, y=ave_price, color=class_gdp)) + 
   geom_line() 
```

seeing this chart we can conclude that GDP isn't necessarily correlated with GDP.
We can see that in high class countries there is a significant difference in food prices
compare with the other three classes. but as we see the difference between the other three 
classes the third worse class has the second highest average prices, followed by the
poorest class having the third highest prices, and the second richest class having the 
lowest prices.



Now lets see how food prices in different clusters 
are affected by different variables depending on their agriculture percent to GDP

```{r}
databaseagri<-databasenew %>% 
  group_by(year,class_gdp) %>% 
  summarize(ave_agri_per=ave(agr_gdp_per))

databaseagri<-databaseagri %>% distinct()

database04new<-inner_join(database02new,databaseagri,by=c("year","class_gdp"))

database04new<-database04new %>% select(1,2,4,5)

ggplot(database04new, aes(x=year, y=ave_agri_per, color=class_gdp)) + 
   geom_line() 

```

```{r}
databaseppp<-databasenew %>% 
  group_by(year,class_gdp) %>% 
  summarize(ave_ppp=ave(purchasingpp))

databaseappp<-databaseppp %>% distinct()

database04new<-inner_join(database04new,databaseppp,by=c("year","class_gdp"))

database04new<-database04new %>% distinct()

ggplot(database04new, aes(x=year, y=ave_ppp, color=class_gdp)) + 
   geom_line() 
```


```{r}
databaseoil<-inner_join(databasenew,oil_prices,by=c("year","country"))

databaseoil2<-databaseoil %>% select(4,5,8,12,13)

databaseoil3<-databaseoil2 %>% 
  group_by(year,class_gdp) %>% 
  summarize(ave_oil_p=ave (oil_price))


database05new<-inner_join(database04new,databaseoil3,by=c("year","class_gdp"))

database05new<-database05new %>% distinct()

ggplot(database05new, aes(x=year, y=ave_oil_p, color=class_gdp)) + 
   geom_line() 
```





Adding more variables

Export of goods and services (% of GDP)

```{r}
exp_gdp_perc <- read.csv("~/Desktop/Project/Food Prices/Databases/R-Data/API_NE/exp_gdp_perc.csv")
```

Cleaning database



```{r}
exp_gdp_perc2<-exp_gdp_perc %>% select(1,5:65)

colnames(exp_gdp_perc2)[1] <- "country"

exp_gdp_perc3 <- gather(exp_gdp_perc2,year,exp_perc,-country)
 
exp_gdp_perc3$year <- strtoi(substring(exp_gdp_perc3$year,2))

exp_gdp_perc3[is.na(exp_gdp_perc3)] <- 0
  
exp_gdp_perc3<-exp_gdp_perc3 %>% filter(exp_perc>0)
```


```{r}
database06new<-inner_join(databasenew,exp_gdp_perc3,by=c("year","country"))

database07new<-database06new %>% select(4,12,13)

database07new<-database07new %>% distinct()

database08new<-database07new %>% 
  group_by(year,class_gdp) %>% 
   summarize(median_exp=median(exp_perc))



database04new<-inner_join(database04new,database08new,by=c("year","class_gdp"))


```






```{r}
kwh_per_capita <- read.csv("~/Desktop/Project/Food Prices/Databases/R-Data/API_LP/kwh_per_capita.csv")
```

```{r}
kwh2<-kwh_per_capita %>% select(1,5:65)

colnames(kwh2)[1] <- "country"

kwh2 <- gather(kwh2,year,kwh,-country)
 
kwh2$year <- strtoi(substring(kwh2$year,2))

kwh2[is.na(kwh2)] <- 0
  
kwh2<-kwh2 %>% filter(kwh>0)
```


```{r}
database12new<-inner_join(databasenew,kwh2,by=c("year","country"))

database13new<-database12new %>% select(4,12,13)

database13new<-database13new %>% distinct()

database14new<-database13new %>% 
  group_by(year,class_gdp) %>% 
   summarize(median_kwh=median(kwh))



database04new<-inner_join(database04new,database14new,by=c("year","class_gdp"))

database04new<-database04new %>% distinct()
```






```{r}
agri_land <- read.csv("~/Desktop/Project/Food Prices/Databases/R-Data/API_LP/agri_land.csv")
```




```{r}
agri_land2<-agri_land %>% select(1,5:65)

colnames(agri_land2)[1] <- "country"

agri_land2 <- gather(agri_land2,year,al,-country)
 
agri_land2$year <- strtoi(substring(agri_land2$year,2))

agri_land2[is.na(agri_land2)] <- 0
  
agri_land2<-agri_land2 %>% filter(al>0)
```


```{r}
database15new<-inner_join(databasenew,agri_land2,by=c("year","country"))

database16new<-database15new %>% select(4,12,13)

database16new<-database16new %>% distinct()

database17new<-database16new %>% 
  group_by(year,class_gdp) %>% 
   summarize(ave_al=ave(al))



database04new<-inner_join(database04new,database17new,by=c("year","class_gdp"))

database04new<-database04new %>% distinct()
```



```{r}
inc_share_lowesr_20 <- read.csv("~/Desktop/Project/Food Prices/Databases/R-Data/API_LP/inc_share_lowesr_20.csv")
```


```{r}
inc_share_lowesr_202<-inc_share_lowesr_20 %>% select(1,5:65)

colnames(inc_share_lowesr_202)[1] <- "country"

inc_share_lowesr_202 <- gather(inc_share_lowesr_202,year,is,-country)
 
inc_share_lowesr_202$year <- strtoi(substring(inc_share_lowesr_202$year,2))

inc_share_lowesr_202[is.na(inc_share_lowesr_202)] <- 0
  
inc_share_lowesr_202<-inc_share_lowesr_202 %>% filter(is>0)
```


```{r}
database18new<-inner_join(databasenew,inc_share_lowesr_202,by=c("year","country"))

database19new<-database18new %>% select(4,12,13)

database19new<-database19new %>% distinct()

database20new<-database19new %>% 
  group_by(year,class_gdp) %>% 
   summarize(median_is=median(is))



database04new<-inner_join(database04new,database20new,by=c("year","class_gdp"))

database04new<-database04new %>% distinct()
```






```{r}
Foodlm26 = lm(ave_price ~  ave_ppp + median_exp  + median_kwh + median_is, data = database04new)

summary(Foodlm26)
```

```{r}
gdp_per_capita <- read.csv("~/Desktop/Project/Food Prices/Databases/R-Data/gdp_per_capita.csv")
```


```{r}
gdp_per_capita2<- gdp_per_capita %>% select(1,5:65)

colnames(gdp_per_capita2)[1] <- "country"

gdp_per_capita2 <- gather(gdp_per_capita2,year,gdp,-country)
 
gdp_per_capita2$year <- strtoi(substring(gdp_per_capita2$year,2))

gdp_per_capita2[is.na(gdp_per_capita2)] <- 0
  
gdp_per_capita2<-gdp_per_capita2 %>% filter(gdp>0)
```


```{r}
database21new<-inner_join(databasenew,gdp_per_capita2,by=c("year","country"))

database22new<-database21new %>% select(4,12,13)

database22new<-database22new %>% distinct()

database22new<-database22new %>% 
  group_by(year,class_gdp) %>% 
   summarize(median_gdp=ave(gdp))



database04new<-inner_join(database04new,database22new,by=c("year","class_gdp"))

database04new<-database04new %>% distinct()
```



```{r}
Foodlm26 = lm(ave_price ~  ave_ppp + median_exp  + median_kwh + median_is , data = database04new)

summary(Foodlm26)
```

Final regression formula= Y= 1.8332949+ 0.0005717(ave_ppp)+  
-0.0293854(median_exp)+ 0.0015150 median_kwh + -0.0743372 (median_is) 


##checking residuals

```{r}
tidy(Foodlm26)
augment(Foodlm26)
augment(Foodlm26) %>%  ggplot(aes(x = ave_price, y = .resid))+
   geom_point() + geom_hline(aes(yintercept = 0))
```


```{r}
par(mfrow=c(1,2)) 
hist(Foodlm26$resi) 
qqnorm(Foodlm26$resi) 
qqline(Foodlm26$resi)
```
The histogram is skewed to the left. There are gaps, but nothing to be worried about

The dots in the Q-Q plot appears close to the reference line.

## Test of normality

```{r}
shapiro.test(Foodlm26$residuals)
```


Since P-value is 0.0724 we accept the null hypothesis and believe the residuals
are normally distributed

##Test of linearity

```{r}
food.lof2<-lm(ave_price ~ as.factor (ave_ppp + median_exp  + median_kwh + median_is), data=database04new)
anova(Foodlm26, food.lof2)
```

At a P-value of 5.15 we accept the null hypothesis and conclude the model fits the data


```{r}
databasenewnew<-full_join(databasenew,inc_share_lowesr_202,by=c("year","country"))

databasenewnew2<-full_join(databasenewnew,agri_land2,by=c("year","country"))

databasenewnew4<-full_join(databasenewnew2,kwh2,by=c("year","country"))

databasenewnew5<-full_join(databasenewnew4,exp_gdp_perc3,by=c("year","country"))

```



```{r}
fulldataset<-databasenewnew5 %>% select(1,3,4,5,7,8,9,10,13,14,15,16)

fulldataset<-fulldataset %>% distinct()

skim(fulldataset)
```

